Behavior Driven Development in Haskell

### 2012-05-31 paper.tex
Fix semantic
Add TBD
Many fixes and wording improvements

@@

\begin{itemize}
\item
    For user documentation,
    Haddock does not have markups to describe evidence.
\item
    For code documentation,
    Hspec does not provide combinators to express expectations with
    BDD terminology.
\end{itemize}

@@

As we explained in Section \ref{sec:problem},
we use Doctest for user documentation and Hspec for code documentation.
The following subsections explain two new markups of Haddock
and evidence combinators for Hspec.

@@

\noindent First, we introduce new Haddock markup for examples.
In this paper, we use {\tt >>>} which is the same as Python's one.
Here is an example:

@@

%%%%%%%%%%%%%%%%
\subsubsection{Markup for properties}
\label{sec:markups}

@@

\noindent Of course, we allow quantified properties by using a
lambda expression. The property example in Section~\ref{sec:doctest-haddock}
can be written as follows:

@@

\noindent Note that the $\hookleftarrow$ symbol denotes that two lines are actually written in a single line. If a default type is not suitable, expression type-signatures can be specified for each parameter:

@@

We propose several combinators which can be used to express expectations
about the outcome of examples.  These are built on top of HUnit.

@@

\noindent A common expectation is, that a value should be equal to an other
value.

@@

\noindent Examples are often used to provide evidence for {\tt IO}
actions.

@@

\noindent So we propose a shortcut for this common use case.

@@

\noindent It takes an {\tt IO} action, and a predicate.  The predicate
servers two purposes.  It selects the type of the expected exception,
and constraints the value.

@@

\noindent To make it more convenient, to expect any exception of a
certain type, we provide type-restricted versions of {\tt (const
True)} for all standard exceptions.

@@

We make creative use GHC's mechanism for custom preprocessors for
that.  With our solution, the developer creates test driver, that only
contains a single line:

@@

This tells GHC to run a preprocessor, \verb|hspec-discover| on the
source file.  \verb|hspec-discover| finds all specs for a project, and
generates a suitable test driver.

This approach relies on the following conventions
\begin{enumerate}
    \item
        Spec files have to be placed into the same directory as the
        test driver, or into a subdirectory.
    \item
        The name of a spec file has to end in \verb|Spec.hs|, and the
        module name has to match the file name.
    \item
        Each spec file has to provide a top-level binding \verb|spec|,
        of type \verb|Spec|.
\end{enumerate}

It would be possible to lift some of those requirements.  But we think
they are reasonable conventions, anyway.

@@

Haddock version 2.7.2 or earlier provided markup syntax for
code blocks, but did not provide any markup syntax for evidence.
As we described, we implemented markup support for examples ({\tt >>>}) in 2010.
It is available in Haddock version 2.8 or later.
We also extended Haddock to support markup for properties ({\tt prop>}) which
will be available with a future release of Haddock.
Both examples and properties in Haddock comments are pretty-printed when
generating documentation, (e.g. as HTML).

@@

To extract evidence from Haddock markup,
Doctest used Haddock API in the past.
Since Each release of Haddock supports only the latest version of GHC,
it is difficult to support old versions of GHC with Haddock API.
Doctest is now directly using GHC API for portability.

@@

Doctest reuses the same GHCi session for all examples.
Examples from the same Haddock comment are grouped together,
and share the same scope.
The scope is cleared with \verb|:reload| before a new
Example group is started.

@@

\noindent Here is the example in Section~\ref{sec:markups} again:

@@

\noindent All combinators introduced in \ref{sec:code-doc} have been
implemented in the {\tt hspec-shouldbe} package\footnote{%
http://hackage.haskell.org/package/hspec-shouldbe};
and they will be incorporated into a future
release of Hspec.
The implementation is straightforward, so we will not further discuss
it.

@@

\newpage % FIXME
\subsection{hspec-discover}
\label{sec:implementation-discovery}

As discussed in \ref{sec:solution-discovery}, we use a preprocessor to
automatically generate test drivers.  GHC invokes preprocessors with
three arguments.

\begin{description}
    \item[{\tt src}]
        -- the original source file\\A preprocessor usually does not
        look at the content of this file.  It is only provided, so
        that it can be used in error messages.
    \item[{\tt cur}]
        -- the file that holds the input for the preprocessor
    \item[{\tt dst}]
        -- a path, where the preprocessor should write it's result to
\end{description}

\noindent A typical preprocessor reads the content of {\tt cur},
processes it, and writes the result to {\tt dst}.  If something goes
wrong, it uses {\tt src} in error messages.

\verb|hspec-discover| ignores {\tt cur} altogether.  It uses the
directory component of {\tt src}, to find all files that are in the
same directory as {\tt src}, or any subdirectories.  It than generates
a test driver from all files that end in {\tt Spec.hs}.

@@

@@

\noindent Now we run test suites until
all test failures disappear.

@@

\begin{quote}
\emph{This is optional, so we can skip it, if necessary.}
\end{quote}

>>>>

\begin{quote}
    TBD: introduce \emph{test driver}
\end{quote}

@@

\begin{itemize}
\item
    For user documentation,
    Haddock does not have markup to describe evidence.
\item
    For code documentation,
    Hspec does not provide combinators to express expectations with
    BDD terminology.
\end{itemize}

@@

As we explained in Section \ref{sec:problem},
we use Doctest for user documentation and Hspec for code documentation.
The following subsections explain two extensions to Haddock's markup.
And combinators for expectations, that can be used with Hspec, are
introduced.

@@

\noindent First, we introduce new Haddock markup for interactive examples.
In this paper, we use {\tt >>>} which is the same as Python's one.
Here is an example:

@@

%%%%%%%%%%%%%%%%
\subsubsection{Markup for properties}
\label{sec:markup}

@@

\noindent And it is possible to quantify properties by using a
lambda expression. The property from Section~\ref{sec:doctest-haddock}
can be written as:

@@

\noindent Note that the $\hookleftarrow$ symbol denotes line
continuation.
If a default type is not suitable, expression type-signatures can be
specified for each parameter:

@@

We propose several combinators which can be used to set expectations
about the outcome of code examples.  These are built on top of HUnit:

@@

\noindent A common expectation is, that a value should be equal to an other
value:

@@

\noindent Code examples are often used to provide evidence for {\tt
IO} actions, e.g.:

@@

\noindent So we propose a shortcut for this common use case:

@@

\noindent It takes an {\tt IO} action, and a predicate.  The predicate
servers two purposes.  It selects the type of the expected exception,
and constraints the value.  Here is an example:

@@

\noindent To make it more convenient to expect any exception of a
certain type, we provide type-restricted versions of {\tt (const
True)} for all standard exceptions, e.g.:

@@

We make creative use GHC's support for custom preprocessors for
that.  With our solution, the developer creates a test driver, that only
contains a single line:

@@

\noindent This tells GHC to run a preprocessor, \verb|hspec-discover|, on the
source file.  \verb|hspec-discover| finds all specs for a project, and
generates a suitable test driver.

This approach relies on the following conventions:
\begin{enumerate}
    \item
        Spec files have to be placed into the same directory as the
        test driver, or into a subdirectory.
    \item
        The name of a spec file has to end in \verb|Spec.hs|, and the
        module name has to match the file name.
    \item
        Each spec file has to export a top-level binding \verb|spec|,
        of type \verb|Spec|.
\end{enumerate}

\noindent It would be possible to lift some of those requirements.
But we think they are reasonable conventions, anyway.

@@

\begin{quote}
    TBD: This section needs more work, to get the wording right.
\end{quote}

Haddock version 2.7.2 and earlier provided markup syntax for
code blocks, but did not provide any markup syntax for evidence.
As we described, we implemented markup support for interactive examples ({\tt >>>}) in 2010.
It is available in Haddock version 2.8 or later.
We also extended Haddock to support markup for properties ({\tt prop>}) which
will be available with a future release of Haddock.
Both interactive examples and properties in Haddock comments are
include into generated documentation.

@@

To extract evidence from Haddock comments,
Doctest used the Haddock API in the past.
Since each release of Haddock supports only the latest version of GHC,
it is difficult to support old versions of GHC with Haddock API.
Doctest is now directly using GHC API for portability.

@@

Doctest reuses the same GHCi session for all examples.
Examples from the same Haddock comment are grouped together,
and share the same scope.
The scope is cleared with \verb|:reload| before a new
example group is started.

@@

\noindent Here is the example from Section~\ref{sec:markup}:

@@

\noindent All combinators introduced in \ref{sec:code-doc} have been
implemented.  They are available form Hackage\footnote{%
http://hackage.haskell.org/package/hspec-shouldbe},
and they will be incorporated into a future
release of Hspec.
The implementation is straightforward, so we will not further discuss
it.

@@

\subsection{hspec-discover}
\label{sec:implementation-discovery}

As discussed in \ref{sec:solution-discovery}, we use a preprocessor to
automatically generate test drivers for Hspec.  GHC invokes the
preprocessor with three arguments:

\begin{description}
    \item[{\tt src}]
        -- path to the original source file\\
        A preprocessor usually does not
        look at the content of this file.  It is only provided, so
        that it can be used in error messages.
    \item[{\tt cur}]
        -- path to a file that holds the input for the preprocessor
    \item[{\tt dst}]
        -- a path, where the preprocessor should write it's result to
\end{description}

\noindent A typical preprocessor reads the content of {\tt cur},
processes it, and writes the result to {\tt dst}.  If something goes
wrong, it uses {\tt src} in error messages to refer to the source
file.

\verb|hspec-discover| ignores {\tt cur} altogether.  It uses the
directory component of {\tt src}, to find all files that are in the
same directory as the source file, or in any subdirectories.  It then generates
a test driver from all files that end in {\tt Spec.hs}.

@@

\begin{quote}
    sol: TBD. @kazu, the descriptions make no sense, can you try to
    make them more precise?
\end{quote}

@@

\noindent Now we run tests, and improve our implementation, until all
tests pass.

@@
